---
title: "Session 4 - Data Visalization 1"
author: "R Introduction Workshop"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
      html_document:   
       toc: true
       toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message=F, eval=T, fig.align = "center")
library(tidyverse)
library(lattice)
library(plotly)

### IMPORTANT !!! ###
# change the path to your directory
setwd("~/Google Drive/RLadiesColumbus/RIntro/Sessions/Session 3/")
```

# Visualization with R

- R has a variety of visualization tools.   
- These tools can be general and applied to any field or data.   
- Some packages can also be very field specific.  

**There are 3 main families of visualization functions:**    

- Base R ` plot() ` -- See `?plot`   
- Lattice -- See `?lattice::Lattice`   
- ggplot2 -- See `?ggplot2::ggplot2`     
- Let's check out [**The R Graph Gallery**](http://www.r-graph-gallery.com/)


# Base R visualization

 - Simplest plotting in R   
 - Not very pretty
 - Difficult to customize   
 - Not so bad if you just want to use it for quick data exploration

**Basic plot syntax:**   

`plot(x , y)` `x`: vector for x axis, `y`: vector for y axis    

See `?plot`  

```{r, eval=T, echo=T, fig.align='center'}
x <- 1:10 
y <- 1:10
plot(x, y)
```


## Base R visualization: Scatterplot  with `iris`

```{r, fig.align='center'}
plot(iris$Sepal.Width, iris$Sepal.Length)
```

## Base R visualization: Histogram with `iris`

```{r, fig.align='center'}
hist(iris$Sepal.Width)
```

## Base R visualization: Using `par()` to plot multiple plots

```{r, fig.align='center', out.width='90%', fig.width=9}
par(mfrow=c(1,2))
plot(iris$Sepal.Width, iris$Sepal.Length)
hist(iris$Sepal.Width)
```

  
# `plot()` vs `ggplot()`  

> - A picture is worth a thousand words -- when the picture is good

```{r, echo=F, fig.align='default', fig.height=6, fig.show='hold', out.width = '50%', message=F}
plot(iris$Sepal.Width, iris$Sepal.Length)
#dotplot(iris$Sepal.Width, iris$Sepal.Length)
qplot(Sepal.Width, Sepal.Length, data = iris, color = Species)
```

## Add layers to `ggplot()`

```{r, echo=F,fig.align='default', fig.height=6, fig.show='hold', out.width = '50%', message=F}
qplot(Sepal.Width, Sepal.Length, data = iris, color = Species)
qplot(Sepal.Width, Sepal.Length, data = iris, color = Species, shape=Species, geom=c("point", "smooth"), main="ggplot")
```

## And make it interactive with `ggplotly()`

```{r, echo=F, message=F, fig.height=8, fig.width=10}
qiris <- qplot(Sepal.Width, Sepal.Length, data = iris, color = Species, shape=Species, geom=c("point", "smooth"), main="ggplot")
ggplotly(qiris)
```

# ggplot2: Create Elegant Data Visualisations Using the Grammar of Graphics

> - A system for 'declaratively' creating graphics, based on "The Grammar of Graphics". 
> - You provide the data, tell `ggplot2` how to map variables to aesthetics, what graphical primitives to use, and it takes care of the details.
> - Infinite options for the ultimate customization
> - It is part of the [tidyverse](http://tidyverse.org/), a collection of R packages that share common philosophies and are designed to work together.

## Installation

```{r, eval=F}
# The easiest way to get ggplot2 is to install the whole tidyverse:
install.packages("tidyverse")

# Alternatively, install just ggplot2:
install.packages("ggplot2")

# Don't forget to load tidyverse to your environment
library(tidyverse)

# Or just ggplot2
library(ggplot2)
```

## Usage
   
1. Start with `ggplot()`, 
    - supply a dataset   
    - and aesthetic mapping using `aes()`. 
    
2. You can then add on layers such as: 
    - Geom (geometric object) with various `geom_` functions.
    - Scales with various `scale_` or `labs()` and `lims()` functions.
    - Faceting specifications with `facet_` functions   
    - Coordinate systems with `coord_` functions   

# Building a ggplot from scratch with `iris`   

## Step 0: Let's remember the `iris` data 
```{r}
head(iris, 3)
summary(iris)
```


## Step 1. Define data and aesthetics woth `aes()`

```{r}
p <- ggplot(data=iris, aes(x=Sepal.Width, y=Sepal.Length))
p
```

## Step 2. Define plot type with `geom_`

```{r}
p <- ggplot(data=iris, aes(x=Sepal.Width, y=Sepal.Length))
p + geom_point()
```

## Step 3. Assign more aesthetics

### Step 3.1 Add `color`

```{r}
p <- ggplot(data=iris, aes(x=Sepal.Width, y=Sepal.Length))
p + geom_point(aes(color=Species))
```


### Step 3.2 Add `color` + `size`

```{r}
p <- ggplot(data=iris, aes(x=Sepal.Width, y=Sepal.Length))
p + geom_point(aes(color=Species, size=Petal.Length))
```

### Step 3.3  Add `color` + `size` + `alpha` (transparency)

```{r}
p <- ggplot(data=iris, aes(x=Sepal.Width, y=Sepal.Length))
p + geom_point(aes(color=Species, size=Petal.Length, alpha=Petal.Width))
```

### Step 3.4  Add `color` + `size` + `alpha` + `shape`

```{r}
p <- ggplot(data=iris, aes(x=Sepal.Width, y=Sepal.Length))
p + geom_point(aes(color=Species, size=Petal.Length, alpha=Petal.Width, shape=Species))
```

## Step 4. Customize legend

```{r}
p <- ggplot(data=iris, aes(x=Sepal.Width, y=Sepal.Length))
p + geom_point(aes(color=Species, size=Petal.Length, alpha=Petal.Width, shape=Species)) +
    guides( color=guide_legend(ncol = 3, byrow = TRUE), 
            size=guide_legend(ncol = 3, byrow = TRUE), 
            alpha=guide_legend(ncol = 3, byrow = TRUE))
```

## Step 5: Assign more `geom`: `point` + `smooth` 

```{r, eval=F}
p <- ggplot(data=iris, aes(x=Sepal.Width, y=Sepal.Length))
p + geom_point(aes(color=Species, size=Petal.Length, alpha=Petal.Width)) + 
    geom_smooth()
```

What will this give me?


```{r, eval=T, echo=F}
p <- ggplot(data=iris, aes(x=Sepal.Width, y=Sepal.Length))
p + geom_point(aes(color=Species, size=Petal.Length, alpha=Petal.Width)) + 
    geom_smooth()
```

Ooops! What happened??  

- Smooth line was for the entire data and **not for each of the Iris species.** 
- Why?

### Step 5.1: Assign more `geom`: `point` + `smooth`

```{r, eval=T}
p <- ggplot(data=iris, aes(x=Sepal.Width, y=Sepal.Length, color=Species))
p + geom_point(aes(size=Petal.Length, alpha=Petal.Width)) + 
    geom_smooth()
```

Why did this work now?   
Can you see the difference?

### Step 5.2: Assign more `geom`: `point` + `smooth`

```{r,eval=T, fig.align='center'}
p <- ggplot(data=iris, aes(x=Sepal.Width, y=Sepal.Length))
p + geom_point(aes(size=Petal.Length, alpha=Petal.Width)) + geom_smooth(aes(color=Species))
```

What about this?
What's happening here?

Now the color is only defined in the `geom_smooth` and not for `geom_point`

## Step 6: Facetting

### Step 6.0: Create a toy dataset

Let's generate a hypothetical `iris` with some added ecosystem type and precipitation data.

- 3 types of ecosystem: Forest, Riparian, Urban
- 2 types of precipitation: Heavy, Mild

Here I am using the `sample()` function which allows me to randomly pick variables from a vector.   
I assign the 150 variables I want to pick with `size = 150` and `replace=TRUE` allows me to pich these variables multiple times.   
Then I assign these to my new dataset `iris2` with column names `Ecosystem` and `Precipitation`

```{r}
ecosys <- sample(c("Forest", "Riparian", "Urban"), size = 150, replace = T)
precp <- sample(c("Heavy", "Mild"), size = 150, replace = T)

iris2 <- cbind(iris, Ecosystem=ecosys, Precipitation=precp)
head(iris2)
```


### Step 6.1: Facet `iris2`

Now, I would like to see how my previous graph changes for the different types of ecosystem and precipitation.    

This was the graph :    

- I am not using `geom_smooth` for now because I do not have enough data points for model prediction.    
- Also I will remove the `alpha` aesthetic to make it easier for us to see.

```{r}
p2 <- ggplot(data=iris2, aes(x=Sepal.Width, y=Sepal.Length, color=Species))
p2 <- p2 + geom_point(aes(size=Petal.Length)) # + geom_smooth() 
p2
```

**Now I add facets!**

```{r}
p2 + facet_grid(Ecosystem ~ Precipitation) 
```

**I can customize the facets very easily!**

```{r}
p2 + facet_grid( . ~ Precipitation) 
```

```{r}
p2 + facet_grid(Ecosystem ~ .) 
```

```{r}
p2 + facet_grid(Precipitation ~ .) 
```

You get the idea here right?

### Step 6.2: Facet `wages`

You can use `facet_wrap` if you want to facet by just 1 variable but you want to organize them nicely.

First, let's read the `wages` data in R.    

```{r}
wages <- read.csv("./wages.csv", 
                   header = T, 
                   stringsAsFactors = T) 
head(wages, 3)
```

Let's create age categories with `cut()` function. I am turning the continous `age` varialbe into a categorical variable with `cut()` function. I am also setting categorical intervals increasing by 10  years with the `breaks = seq(20, 100, by=10)` argument. Then I assign this new variable to a new column called `age_cat`.   

```{r}
wages <- wages %>% mutate(age_cat = cut(age, breaks = seq(20, 100, by=10)) )
head(wages, 4)
```

Let's plot it

```{r}
pw <- ggplot(wages, aes(x=height, y=earn)) +
      geom_point(aes(size=ed), alpha=0.5)
pw
pw + facet_wrap(~age_cat)
```

Or you can specify the rows and columns for the faceting

```{r}
pw + facet_wrap(~age_cat, ncol=5)
```


# **Your turn** (5 mins)

Plot the `wages.csv` data like the following

```{r, echo=F,fig.width=10, out.width='100%'}
pw <- ggplot(wages, aes(x=age, y=earn, color=sex))
pw + geom_point(aes(size=ed), alpha=0.5) + 
     geom_smooth() +
     facet_wrap(~ race)
```