---
title: "Session 1"
author: "R Introduction Workshop"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    toc: true
---

# Shortcuts

 1. When your cursor is in **console** you can tap on the **up arrow** to bring back your last executed command. If you tap twice you can bring back the second last executed command. If you tap three times you can bring back the third last executed command. See the pattern there...?
 
 2. To complete an existing function or variable tap **Tab** and it will show you the options.
 
 3. To bring back a certain executed command that starts with a certain character (or letter) tap **Ctrl** (for Windows) or **Cmd** (for Mac) at the same time with **up arrow**.
 
 4. In order to execute a command on your R script you click on **Run** top right on your script pane. Or you can tap **Ctrl** (for Windows) or **Cmd** (for Mac) and **Enter** 
 
 5. You can paste your history from the **history pane**, by clicking **To console** (to copy it to the console) and **To source** (to copy it to the scrip pane).
 
# R Objects

## Assigning R objects

Save information as an R object with the greater than sign followed by a minus, e.g, an arrow: `<-`

```{r}
foo <- 42
foo + 12
```

When you create an R object, you'll see it appear in your **Environment pane**

## Common R workflow

Usually in R you would assign the output of a function to a variable to save it and then to use it in a second function.

```{r}
foo <- round(3.1415) + 1 
foo
factorial(foo)
```

Because if you don't save the output in a variable, you would only output it to the console. And can't use it as an input for the next function.

```{r}
round(3.1415) + 1 
```

## Object names

 1. Object names cannot begin with numbers, or R special characters.
 These as variable names will give you errors: `1st`, `$income`, `>data`, `!bad`, `^mean` 
 
 2. Be careful when reassigning names already in use.
 
 3. Capitalization matters. R will treat each of these as a different object:
 `MyData` vs `mydata`
 `A` vs `a`
 `sum` (a R function) vs `SUM`
 
 4. You can remove an object (including the data contained in it) with `rm` function.
 Simply put the name of the variable in the `rm` e.g. `rm(foo)`
 

# Data Structures

## Vectors
Concatenate multiple elements into a one dimensional array.
Create with the `c()` function.

```{r}
vec <- c(1, 2, 3, 10, 100)
vec
```

## Matrices

Multiple elements stored in a two dimensional array.
Create with the `matrix()` function

```{r}
mat <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 2)
mat
```

You can also transpose a matrix with `t()` function.

```{r}
t(mat)
```


## Math: element-wise

```{r}
vec
vec + 4

vec * 4

vec * vec
```

# Data types (classes)

Like Excel, Numbers, etc., R can recognize different types of data.
We'll look at four basic types:

 * numbers     
 * character strings (text)   
 * logical   
 * factor
 
Check the type (class) of any R object with `class()` function 
 
## Numeric

Any number, no quotes.
Appropriate for math.
 
```{r}
1+ 1
3000000
class(0.00001)
```

## Character

Any symbols surrounded by quotes.
Appropriate for words, variable names, messages, any text.

```{r}
"hello"
class("hello")
```

Check the number of characters in a character element.
```{r}
nchar("hello")
```

Merge two character elements into one with `paste()`

```{r}
paste("hello", "world")
```

## Logical

`TRUE` (or just `T`) or `FALSE` (or just `F`)
R's form of binary data. Useful for logical tests.

```{r}
3< 4
class(TRUE)
class(T)
class(FALSE)
class(F)
```

## Factor

R's form of categorical data. Saved as an integer with a set of labels (e.g. levels).

```{r}
vec2 <- c("a", "b","b", "c", "d", "a") # vec2 is a character vector
fac <- factor(vec2) #now fac is factor
class(fac)
fac
```

One can easily change the "labels" e.g. `levels` of a factor

Let's change `a` to `apple` and `d` to `dragonfruit`


```{r}
fac
levels(fac)
levels(fac) <- c("apple", "b", "c", "dragonfruit")
fac
```

## Classes and Data Structures

Vector and matrices cannot have different class elements. For example a vector cannot contain a character, a number or a logical all at the same time. Therefore if there are **any** characters, all other elements of the vector are coerced to character class. 

```{r}
v1 <- c("R", 1, TRUE)
class(v1)
v1
```

If the vector contains numeric and logical elements, all will coerce to numeric, where `TRUE` equals to `1` and `FALSE` equals to `0`.

```{r}
v2 <- c(1, TRUE)
class(v2)
v2
```

**What if you want different data types in the same object?**

List and data frames generalize vectors and matrices to allow multiple types of data.

## List
A list is a one dimensional group of R objects.
Create lists with `list()`

```{r}
lst <- list(1, "R", TRUE) 
class(lst)
lst
```

The elements of a list can be anything. Even vectors, matrices, data frames or other lists. Trippy isn't it?

```{r}
list(c(1, 2), TRUE, c("a", "b", "c"), matrix(c(1:10), nrow = 2))
```

## Data frame

A data frame is a two dimensional group of R objects.
Each column in a data frame can be a different type (class).
(But the elements of a column cannot be of different type.)

```{r}
df <- data.frame( c(1, 2, 3),  #this one is numeric
                  c("R","S","T"), # this one is character
                  c(TRUE, FALSE, TRUE)) #this one is logical
class(df)
df
```

## Names

You can name the elements of a vector, list, or data frame when you create them.

```{r}
#vector
nvec <- c(one = 1, two = 2, three = 3)
nvec

# list
nlst <- list(one = 1, 
             two = 2,
             many = c(3, 4, 5))
 
nlst

# data frame
ndf <- data.frame(numbers = c(1, 2, 3),
                  letters = c("R","S","T"),
                  logic = c(TRUE, FALSE, TRUE))
```

You can also see and set the names with `names()`. (Similar to `levels()`, do you remember that one?)

```{r}
names(nvec)
names(nvec) <- c("uno", "dos", "tres") 
nvec
```

Similarly you can change the name of rows and columns of a data frame as well.

```{r}
colnames(ndf)
colnames(ndf) <- c("nombres", "lettres", "logique") # yes, it's french
ndf

rownames(ndf)
rownames(ndf) <- c("myRow1", "myRow2", "myRow3")
ndf
```



Once you have the names set up, you can call columns of a data frame or elements of a vector or a list very easily.

```{r}
# vector
nvec["uno"]

# list -- attention to double brackets
nlst[["many"]]
# or another way
nlst$many

# data frame
ndf["nombres"]
# or another way
ndf$nombres
```

Can you tell the difference between `ndf["numbers"]` and `ndf$numbers`?   
They're different type of data structures!    
`ndf["nombres"]` is a data frame with only one column, `ndf$nombres` is a numeric vector.

```{r}
class(ndf["nombres"])
class(ndf$nombres)
```



